Concentration of Maxima and Fundamental Limits in High-Dimensional Testing and Inference

### 2021-02-04 5.Bayes.optimality/5.Bayes_optimality_and_minimax.tex
remove "Proof (Proof of"

\begin{proof}[Proof of Corollary \ref{cor:log-convex}]
Since the density of the alternatives $f_{a}(t) = {f(t-\delta)}$ is log-convex on $[K+\delta, \infty)$, by Relation \eqref{eq:concavity-implies-MLR} in the proof of Lemma \ref{lemma:MLR-log-concavity} and appealing to log-convexity (rather than log-concavity), the likelihood ratio
$$
L(j) := \frac{f_a(x(j))}{f_0(x(j))} %= \sum_{i=1}^s \frac{p_if_{i}(x(j))}{f_0(x(j))},
$$
is decreasing in $x(j)$ on $[K+\delta, \infty)$.  The claim follows from Proposition \ref{prop:likelihood-ratio-thresholding}.
\end{proof}

>>>>

\begin{proof}[Corollary \ref{cor:log-convex}]
Since the density of the alternatives $f_{a}(t) = {f(t-\delta)}$ is log-convex on $[K+\delta, \infty)$, by Relation \eqref{eq:concavity-implies-MLR} in the proof of Lemma \ref{lemma:MLR-log-concavity} and appealing to log-convexity (rather than log-concavity), the likelihood ratio
$$
L(j) := \frac{f_a(x(j))}{f_0(x(j))} %= \sum_{i=1}^s \frac{p_if_{i}(x(j))}{f_0(x(j))},
$$
is decreasing in $x(j)$ on $[K+\delta, \infty)$.  The claim follows from Proposition \ref{prop:likelihood-ratio-thresholding}.
\end{proof}

### 2021-02-04 6.GWAS/6.proofs-GWAS.tex
remove "Proof (Proof of"

We review some properties of the chi-square distributions in Section \ref{sec:chi-square-distributions}, before presenting the proofs of the main theorems on phase transitions in Sections \ref{subsec:proof-chi-squared-exact-boundary}, \ref{subsec:proof-chi-squared-approx-boundary}, and \ref{subsec:proof-chi-squared-mix-boundaries}.
Results relating signal sizes and effect sizes in association tests will be justified in Section \ref{subsec:proof-signal-size-odds-ratio}.

@@

\begin{proof}[Proof of Theorem \ref{thm:chi-squared-exact-boundary}]
We first prove the sufficient condition.
The Bonferroni procedure sets the threshold at $t_p = F^\leftarrow(1-\alpha/p)$, which, by Lemma \ref{lemma:chisq-quantiles}, is asymptotic to $2\log{p} - 2\log{\alpha}$.
By the assumption on $\alpha$ in \eqref{eq:slowly-vanishing-error}, for any $\delta>0$, we have $p^{-\delta}=o(\alpha)$.
Therefore, we have $-\log\alpha\le\delta\log{p}$ for large $p$, and
\begin{equation*} 
    1 \le \limsup_{p\to\infty}\frac{2\log{p} - 2\log{\alpha}}{2\log{p}} \le 1+\delta,
\end{equation*}
for any $\delta>0$.
Hence, $t_p\sim 2\log{p}$.

@@

\begin{proof}[Proof of necessary condition in Theorem \ref{thm:chi-squared-approx-boundary}]
Denote the distributions of $\chi^2_\nu(0)$, $\chi^2_\nu(\underline{\Delta})$ and $\chi^2_\nu(\overline{\Delta})$ as $F_0$, $F_{\underline{a}}$, and $F_{\overline{a}}$ respectively.

@@

\begin{proof}[Proof of sufficient condition in Theorem \ref{thm:chi-squared-approx-boundary}]
The FDR vanishes by our choice of $\alpha$ and the FDR-controlling property of the BH procedure.
It only remains to show that FNR also vanishes.

@@

\begin{proof}[Proof of Theorem \ref{thm:chi-squared-exact-approx-boundary}]
% The reduction to equal signal sizes can be achieved 
We first show the sufficient condition.
Similar to the proof of Theorem \ref{thm:chi-squared-approx-boundary}, it suffices to show that
\begin{equation} \label{eq:exact-approx-boundary-proof-sufficient-1}
    \text{NDP} = 1 - \widehat{W}_\text{signal}(t_p) \to 0,
\end{equation}
where $t_p$ is the threshold of Bonferroni's procedure.

@@

\begin{proof}[Proof of Theorem \ref{thm:chi-squared-approx-exact-boundary}]
We first show the sufficient condition.
Since FDR control is guaranteed by the BH procedure, we only need to show that the FWNR also vanishes, that is,
\begin{equation} \label{eq:approx-exact-boundary-proof-sufficient-1}
    \P\left[\min_{i\in S}x(i) \ge \tau\right] \to 1,
\end{equation}
where $\tau$ is the threshold for the BH procedure.

@@

\begin{proof}[Proof of Lemma \ref{lemma:relative-stability-order-statistics}]
Using the Renyi representation for order statistics, we write
\begin{equation} \label{eq:relative-stability-order-statistics-proof-1}
    Z_{[i]} = \Phi^{\leftarrow}(U_{[i]}),
\end{equation}
where $U_{[i]}$ is the $i^\mathrm{th}$ (smallest) order statistic of $s$ independent uniform random variables over $(0,1)$.
Since $U_{[i]}$ has a $\mathrm{Beta}(i, s+1-i)$ distribution, with mean and standard deviation,
$$
\E[U_{[k]}] = k/(s+1) \sim s^{-\beta_0}, 
\quad \text{and} \quad
{\mathrm{sd}(U_{[k]})} = \frac{1}{s+1}\sqrt{\frac{k(s+1-k)}{s+2}} \sim s^{-\frac{1+\beta_0}{2}},
$$
we obtain by Chebyshev's inequality 
% (since $\E[U_{[k]}]/\mathrm{sd}(U_{[k]})\sim p^{(1-\beta_0)/2}$) 
$$
\P\left[s^{-\beta_0}(1-\epsilon) < U_{[k]} < s^{-\beta_0}(1+\epsilon)\right] \to 1,
$$
where $\epsilon$ is an arbitrary positive constant.
This implies, by representation \eqref{eq:relative-stability-order-statistics-proof-1},
\begin{equation} \label{eq:relative-stability-order-statistics-proof-2}
    \P\left[\Phi^{\leftarrow}\left(s^{-\beta_0}(1-\epsilon)\right) < Z_{[k]} < \Phi^{\leftarrow}\left(s^{-\beta_0}(1+\epsilon)\right)\right] \to 1.
\end{equation}
Using the expression for standard Gaussian quantiles (see, e.g., Proposition 1.1. in \cite{gao2018fundamental}), we know that
\begin{align*}
    \Phi^{\leftarrow}\left(s^{-\beta_0}(1-\epsilon)\right) 
    &\sim -\sqrt{2\log{\left(s^{\beta_0}/(1-\epsilon)\right)}} \\
    &= -\sqrt{2(\beta_0\log{s} - \log{(1-\epsilon)})} \sim -\sqrt{2\beta_0\log{s}},
\end{align*}
and similarly $\Phi^{\leftarrow}\left(s^{-\beta_0}(1+\epsilon)\right)\sim -\sqrt{2\beta_0\log{s}}$.
Since both ends of the interval in \eqref{eq:relative-stability-order-statistics-proof-2} are asymptotic to $-\sqrt{2\beta_0\log{s}}$, 
% dividing through by $\sqrt{2\log{s}}$, and 
the desired conclusion follows.
\end{proof}

>>>>

We review some properties of the chi-square distributions in Section \ref{sec:chi-square-distributions}, before presenting the proofs of the main theorems on phase transitions in Sections \ref{subsec:proof-chi-squared-exact-boundary}, \ref{subsec:proof-chi-squared-approx-boundary}, and \ref{subsec:proof-chi-squared-mix-boundaries}.
%Results relating signal sizes and effect sizes in association tests will be justified in Section \ref{subsec:proof-signal-size-odds-ratio}.

@@

\begin{proof}[Theorem \ref{thm:chi-squared-exact-boundary}]
We first prove the sufficient condition.
The Bonferroni procedure sets the threshold at $t_p = F^\leftarrow(1-\alpha/p)$, which, by Lemma \ref{lemma:chisq-quantiles}, is asymptotic to $2\log{p} - 2\log{\alpha}$.
By the assumption on $\alpha$ in \eqref{eq:slowly-vanishing-error}, for any $\delta>0$, we have $p^{-\delta}=o(\alpha)$.
Therefore, we have $-\log\alpha\le\delta\log{p}$ for large $p$, and
\begin{equation*} 
    1 \le \limsup_{p\to\infty}\frac{2\log{p} - 2\log{\alpha}}{2\log{p}} \le 1+\delta,
\end{equation*}
for any $\delta>0$.
Hence, $t_p\sim 2\log{p}$.

@@

\begin{proof}[Necessary condition in Theorem \ref{thm:chi-squared-approx-boundary}]
Denote the distributions of $\chi^2_\nu(0)$, $\chi^2_\nu(\underline{\Delta})$ and $\chi^2_\nu(\overline{\Delta})$ as $F_0$, $F_{\underline{a}}$, and $F_{\overline{a}}$ respectively.

@@

\begin{proof}[Sufficient condition in Theorem \ref{thm:chi-squared-approx-boundary}]
The FDR vanishes by our choice of $\alpha$ and the FDR-controlling property of the BH procedure.
It only remains to show that FNR also vanishes.

@@

\begin{proof}[Theorem \ref{thm:chi-squared-exact-approx-boundary}]
% The reduction to equal signal sizes can be achieved 
We first show the sufficient condition.
Similar to the proof of Theorem \ref{thm:chi-squared-approx-boundary}, it suffices to show that
\begin{equation} \label{eq:exact-approx-boundary-proof-sufficient-1}
    \text{NDP} = 1 - \widehat{W}_\text{signal}(t_p) \to 0,
\end{equation}
where $t_p$ is the threshold of Bonferroni's procedure.

@@

\begin{proof}[Theorem \ref{thm:chi-squared-approx-exact-boundary}]
We first show the sufficient condition.
Since FDR control is guaranteed by the BH procedure, we only need to show that the FWNR also vanishes, that is,
\begin{equation} \label{eq:approx-exact-boundary-proof-sufficient-1}
    \P\left[\min_{i\in S}x(i) \ge \tau\right] \to 1,
\end{equation}
where $\tau$ is the threshold for the BH procedure.

@@

\begin{proof}[Lemma \ref{lemma:relative-stability-order-statistics}]
Using the Renyi representation for order statistics, we write
\begin{equation} \label{eq:relative-stability-order-statistics-proof-1}
    Z_{[i]} = \Phi^{\leftarrow}(U_{[i]}),
\end{equation}
where $U_{[i]}$ is the $i^\mathrm{th}$ (smallest) order statistic of $s$ independent uniform random variables over $(0,1)$.
Since $U_{[i]}$ has a $\mathrm{Beta}(i, s+1-i)$ distribution, with mean and standard deviation,
$$
\E[U_{[k]}] = k/(s+1) \sim s^{-\beta_0}, 
\quad \text{and} \quad
{\mathrm{sd}(U_{[k]})} = \frac{1}{s+1}\sqrt{\frac{k(s+1-k)}{s+2}} \sim s^{-\frac{1+\beta_0}{2}},
$$
we obtain by Chebyshev's inequality 
% (since $\E[U_{[k]}]/\mathrm{sd}(U_{[k]})\sim p^{(1-\beta_0)/2}$) 
$$
\P\left[s^{-\beta_0}(1-\epsilon) < U_{[k]} < s^{-\beta_0}(1+\epsilon)\right] \to 1,
$$
where $\epsilon$ is an arbitrary positive constant.
This implies, by representation \eqref{eq:relative-stability-order-statistics-proof-1},
\begin{equation} \label{eq:relative-stability-order-statistics-proof-2}
    \P\left[\Phi^{\leftarrow}\left(s^{-\beta_0}(1-\epsilon)\right) < Z_{[k]} < \Phi^{\leftarrow}\left(s^{-\beta_0}(1+\epsilon)\right)\right] \to 1.
\end{equation}
Using the expression for standard Gaussian quantiles (see, e.g., Proposition 1.1. in \cite{gao2018fundamental}), we know that
\begin{align*}
    \Phi^{\leftarrow}\left(s^{-\beta_0}(1-\epsilon)\right) 
    &\sim -\sqrt{2\log{\left(s^{\beta_0}/(1-\epsilon)\right)}} \\
    &= -\sqrt{2(\beta_0\log{s} - \log{(1-\epsilon)})} \sim -\sqrt{2\beta_0\log{s}},
\end{align*}
and similarly $\Phi^{\leftarrow}\left(s^{-\beta_0}(1+\epsilon)\right)\sim -\sqrt{2\beta_0\log{s}}$.
Since both ends of the interval in \eqref{eq:relative-stability-order-statistics-proof-2} are asymptotic to $-\sqrt{2\beta_0\log{s}}$, 
% dividing through by $\sqrt{2\log{s}}$, and 
the desired conclusion follows.
\end{proof}
